---
title: Another test post
author: Jake
date: '2020-05-16'
slug: another-test-post
categories: []
tags: []
description: ''
---
---
title: "Untitled"
author: "Jacob Lowell"
date: "5/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE , message = FALSE  , warning = FALSE )

#, fig.width = 8 , fig.height = 5
```


https://airyards.com/baldwin_boxscores.html

```{r  ,echo=FALSE}
library(tidyverse)
options(scipen=999)
library(RSQLite)
library(DBI)
library(cowplot)
library(gt)
library(tidyquant)
library(glue)
library(tidymodels)
library(plotly)

# create the "empty" database
#mydb <- DBI::dbConnect(RSQLite::SQLite(), "Data/pbp_db.sqlite")


mydb <- DBI::dbConnect(RSQLite::SQLite(), "~/Desktop/Football Project/Data/pbp_db.sqlite")

DBI::dbListTables(mydb)


# Open a queryable connection with the database
pbp_db <- tbl(mydb, "pbp_raw_2000-2019")


library(tictoc)
remove(full_set)

play_by_play <- as_tibble(pbp_db %>%  select(everything()))
```





game_id, game_date, game_seconds_remaining, season_type, week,season,

```{r    message  = FALSE}

#   total_runs ,  total_run_pass_series ,total_runs_series ,total_pass_series  , previous_play_series , previous_play ,  total_pass ,

label_wrap <- label_wrap_gen(width = 60)
#source("scripts/Load data - pass and run plays.R")

# 5  filter so that we have only run and pass data ----

pbp_all_rp <- play_by_play    %>%  filter( season_type == "REG") %>% 
  select(    ydsnet  , yards_gained ,shotgun , no_huddle  ,qb_dropback  ,pass_length  , pass_location  ,
                                air_yards ,     run_location ,   run_gap ,      field_goal_result  ,    fg_prob  ,   td_prob , 
                                home_wp , home_wp_post , away_wp_post  , total_home_rush_wpa  ,total_away_rush_wpa,
                                total_home_pass_wpa , total_away_pass_wpa , 
                                # good fields  
                                score_differential_post,
                                drive_how_started,
                                first_down_rush  ,first_down_pass ,first_down_penalty  ,third_down_converted ,fourth_down_converted ,fourth_down_failed ,
                                
                                rush_attempt  , pass_attempt  ,  passer_player_name  ,receiver_player_name   , rusher_player_name ,  penalty_yards  ,
                                replay_or_challenge , replay_or_challenge_result  ,penalty_type  , cp , cpoe, roof_type , game_time_eastern ,
                                drive_inside20   , drive_first_downs ,
                                
                                # drive
                                drive_how_ended ,drive_play_count  ,  drive_yards_penalized  ,  drive_time_of_possession ,  series  , series_success ,
                                success  ,  passer , rusher , receiver , qb_epa ,
             
             first_down , 
                                
                                                 
  play_type ,play_type_nfl , play  , penalty, desc , ep , epa , wp , wpa , down  , yardline_100 , yards_gained , yards ,pass , rush,
  touchdown ,  td_prob , play_type , down, goal_to_go, desc , yardline_100 , yards_gained , td_team, epa, play_id, game_id, home_team, away_team, posteam, posteam_type, game_date, half_seconds_remaining , game_seconds_remaining, game_half , drive, sp, qtr, ydstogo , shotgun, no_huddle, qb_dropback, qb_scramble, pass_location, air_yards, run_location, run_gap ,posteam_timeouts_remaining , defteam_timeouts_remaining ,  ep , total_home_epa, total_away_epa, total_home_rush_epa , total_home_pass_epa , total_away_pass_epa , total_away_rush_epa ,    def_wp , score_differential , posteam_score , defteam_score  , week , season ,   first_down_pass , first_down_rush , third_down_converted , fourth_down_converted  , passer_player_name , receiver_player_name  , rusher_player_name) %>% 
  filter(td_prob > 0 )%>% 
  filter(!is.na(epa), !is.na(posteam), play_type=="no_play" | play_type=="pass" | play_type=="run") %>%
  mutate(
    pass = if_else(str_detect(desc, "( pass)|(sacked)|(scramble)"), 1, 0),
    rush = if_else(str_detect(desc, "(left end)|(left tackle)|(left guard)|(up the middle)|(right guard)|(right tackle)|(right end)") & pass == 0, 1, 0),
    success = ifelse(epa>0, 1 , 0),
    passer_player_name = ifelse(play_type == "no_play" & pass == 1, 
                                str_extract(desc, "(?<=\\s)[A-Z][a-z]*\\.\\s?[A-Z][A-z]+(\\s(I{2,3})|(IV))?(?=\\s((pass)|(sack)|(scramble)))"),
                                passer_player_name),
    receiver_player_name = ifelse(play_type == "no_play" & str_detect(desc, "pass"), 
                                  str_extract(desc, "(?<=to\\s)[A-Z][a-z]*\\.\\s?[A-Z][A-z]+(\\s(I{2,3})|(IV))?"),
                                  receiver_player_name),
    rusher_player_name = ifelse(play_type == "no_play" & rush == 1, 
                                str_extract(desc, "(?<=\\s)[A-Z][a-z]*\\.\\s?[A-Z][A-z]+(\\s(I{2,3})|(IV))?(?=\\s((left end)|(left tackle)|(left guard)|	 (up the middle)|(right guard)|(right tackle)|(right end)))"),
                                rusher_player_name),
    name = ifelse(!is.na(passer_player_name), passer_player_name, rusher_player_name),
    yards_gained=ifelse(play_type=="no_play",NA,yards_gained),
    play=1  , TD_Scored = ifelse(!is.na(td_team) , "Touchdown" , "No Touchdown")  ,
    Scored_TD = case_when(td_team == posteam ~  "Touchdown" ,
                          is.na(td_team)  ~ "Not Touchdown"  , TRUE ~ "Touchdown Other Team")
  )  %>% 
  filter(play_type %in% c("run" , "pass")  | pass==1 | rush==1)  %>% 
  select(
    play_type ,play_type_nfl , play  , penalty, desc , ep , epa , wp , wpa , down  , yardline_100 , yards_gained , yards ,pass , rush,
    pass , td_prob  ,rush, play_type , down , goal_to_go , desc , yardline_100   , yards_gained , td_team , td_prob , epa   ,everything()) %>% select(-game_date) %>% 
   mutate(play_desc = str_glue("Play: {label_wrap(desc)}")) %>%   mutate(
    stick_throw = case_when(
      air_yards < ydstogo ~ "Short of Sticks",
      air_yards == ydstogo ~ "At Stick",
      air_yards > ydstogo ~ "Past Stick",
      TRUE ~ NA_character_
    )
  ) %>%  mutate(
         run = if_else(play_type == "run", 1, 0),
         pass = if_else(play_type == "pass", 1, 0) ) %>% 
  group_by(game_id, posteam) %>%
  mutate(
    total_runs = if_else(play_type == "run",
                         cumsum(run) - 1, cumsum(run)
    ),
    total_pass = if_else(play_type == "pass",
                         cumsum(pass) - 1, cumsum(pass)
    ),
    previous_play = if_else(posteam == lag(posteam),
                            lag(play_type), "First play of Drive"
    ),
    previous_play = if_else(is.na(previous_play),
                            replace_na("First play of Drive"), previous_play
    )
  )    %>%  ungroup() %>% 
  
  # drive play run or pass count 
   group_by(game_id, posteam , drive) %>%
  mutate(
    total_runs_drive = if_else(play_type == "run",
                         cumsum(run) - 1, cumsum(run)
    ),
    total_pass_drive = if_else(play_type == "pass",
                         cumsum(pass) - 1, cumsum(pass)
    ),
    previous_play_drive = if_else(posteam == lag(posteam),
                            lag(play_type), "First play of Drive"
    ),
    previous_play_drive = if_else(is.na(previous_play_drive),
                            replace_na("First play of Drive"), previous_play_drive
    )
  )   %>%  mutate(total_run_pass_drive = total_runs_drive +  total_pass_drive )%>%  ungroup() %>%  mutate_if(is.character , factor )   # %>%   filter( season != 2019)  
                         


```




## Separate Train and Test sets

```{r}

play_by_play_train <- pbp_all_rp  %>%  filter(season != 2019   &  season != 2000 ) 

play_by_play_test <- pbp_all_rp  %>% filter(season == 2019  ) 

remove(play_by_play)

remove(pbp_all_rp)

```



```{r}

# training sets

first_and_goal <-  play_by_play_train %>%  filter(down == 1 & goal_to_go ==1 ) #%>% count(touchdown)

second_and_goal <-  play_by_play_train %>%  filter(down == 2 & goal_to_go ==1 )

third_and_goal <-  play_by_play_train %>%  filter(down == 3 & goal_to_go ==1 )

fourth_and_goal  <-   play_by_play_train %>%  filter(down == 4 & goal_to_go ==1 )




# test sets

first_and_goal_test <-  play_by_play_test  %>%  filter(down == 1 & goal_to_go ==1 ) #%>% count(touchdown)

second_and_goal_test <-  play_by_play_test  %>%  filter(down == 2 & goal_to_go ==1 )

third_and_goal_test <-  play_by_play_test  %>%  filter(down == 3 & goal_to_go ==1 )

fourth_and_goal_test  <-   play_by_play_test %>%  filter(down == 4 & goal_to_go ==1 )




```






```{r}
# function  with two charts together





cow_plot_epa_wpa <- function(set ){
  w <-   set %>%  ggplot(aes(x=wpa)) + geom_histogram() + labs(title = glue::glue("Win Percentage added on {set[1:2,1]}"))   
  e <-   set %>%  ggplot(aes(x=epa)) + geom_histogram() + labs(title = glue::glue("Expected Points added on {set[1:2,1]}"))   
  
  plot_grid(w,e)
}

cow_plot_epa_wpa(set = first_and_goal)
cow_plot_epa_wpa(set = second_and_goal)
cow_plot_epa_wpa(set = third_and_goal)
cow_plot_epa_wpa(set = fourth_and_goal)


```


```{r}
first_and_goal %>% count(season)
second_and_goal %>% count(season)
third_and_goal %>% count(season)
fourth_and_goal %>% count(season)
```






I am using data from the [nflscrapR](https://arxiv.org/abs/1802.00998) package to investigate offensive success on goal to go situations on different downs.  I will perform exploratory data analysis and modeling on regular season game data from 2009-2018, and test the performance on 2019 regular season data.



# Section 1. Initial Analysis



### Figure 1.

```{r}

play_by_play_train  %>%   filter(down >0 & goal_to_go == 1) %>% ggplot(aes( x =epa ,     fill = TD_Scored )) + geom_histogram() + theme_tq() + 
  labs(title = glue("Goal to go - Distribution of expected points added for each down"             )  ,
       x = "Expected Points Added" , y = "Plays"  ,  fill = "") + scale_color_tq() +
scale_y_continuous(labels = scales::comma)  + facet_wrap(~down   , scales = "free" )




play_by_play_test  %>%   filter(down >0 & goal_to_go == 1) %>% ggplot(aes( x =epa ,     fill = TD_Scored )) + geom_histogram() + theme_tq() + 
  labs(title = glue("Goal to go - Distribution of expected points added for each down"             )  ,
       x = "Expected Points Added" , y = "Plays"  ,  fill = "") + scale_color_tq() +
scale_y_continuous(labels = scales::comma)  + facet_wrap(~down   , scales = "free" )

```


If we define success as a play with a positive value for excepted points added (epa), we can see a nice split in good/bad distribution for 3rd and 4th down.  There is a decent split for second down as well.






```{r}
play_by_play_train %>%  filter(goal_to_go ==1) %>% filter(play_type != "no_play") %>%  count(down, touchdown  ) %>%  group_by( down) %>% mutate(`Touchdown Rate` = 
                round(  n / sum(n)  , digits = 3) )%>% ungroup() %>% 
  filter(touchdown ==1) %>% select(down , `Touchdown Rate`) %>% left_join( play_by_play_train  %>%  filter(goal_to_go ==1) %>%  count(down, success  ) %>%  group_by( down) %>% mutate( `Success Rate`=    round (n / sum(n) , digits = 3)) %>% ungroup() %>%
 filter(success ==1)   )  %>% select(-success , - n)  %>% left_join(play_by_play_train    %>%   filter(down >0 & goal_to_go == 1) %>% group_by(down) %>% summarise(`Mean Expected Points Added` =  round(mean(epa) , digits = 2)   , `Median Expected Points Added` = round(median(epa) , digits = 2)) )   %>% gt() %>%
  fmt_percent(columns = vars(`Touchdown Rate` , `Success Rate`) , decimals = 1)  %>% tab_header(
    title = md("Goal to goal Results"),
    subtitle = ""
  ) %>%
  tab_source_note(md("Source: nflfastR"))
```



```{r}
dog <- play_by_play_train %>%  filter(goal_to_go ==1) %>% filter(play_type == "no_play") %>% select(touchdown , everything())
```



```{r}
play_by_play_test %>%  filter(goal_to_go ==1) %>% filter(play_type != "no_play") %>%  count(down, touchdown  ) %>%  group_by( down) %>% mutate(`Touchdown Rate` = 
                round(  n / sum(n)  , digits = 3) )%>% ungroup() %>% 
  filter(touchdown ==1) %>% select(down , `Touchdown Rate`) %>% left_join( play_by_play_test  %>%  filter(goal_to_go ==1) %>%  count(down, success  ) %>%  group_by( down) %>% mutate( `Success Rate`=    round (n / sum(n) , digits = 3)) %>% ungroup() %>%
 filter(success ==1)   )  %>% select(-success , - n)  %>% left_join(play_by_play_test    %>%   filter(down >0 & goal_to_go == 1) %>% group_by(down) %>% summarise(`Mean Expected Points Added` =  round(mean(epa) , digits = 2)   , `Median Expected Points Added` = round(median(epa) , digits = 2)) )   %>% gt() %>%
  fmt_percent(columns = vars(`Touchdown Rate` , `Success Rate`) , decimals = 1)  %>% tab_header(
    title = md("Goal to goal Results"),
    subtitle = ""
  ) %>%
  tab_source_note(md("Source: nflfastR"))
```





##  plot td rate and success rate over time



```{r}
play_by_play_train %>%  filter(goal_to_go ==1) %>% filter(play_type != "no_play") %>%  count(down, touchdown  ) %>%  group_by( down) %>% mutate(`Touchdown Rate` = 
                round(  n / sum(n)  , digits = 3) )%>% ungroup() %>% 
  filter(touchdown ==1) %>% select(down , `Touchdown Rate`) %>% left_join( play_by_play_train  %>%  filter(goal_to_go ==1) %>%  count(down, success  ) %>%  group_by( down) %>% mutate( `Success Rate`=    round (n / sum(n) , digits = 3)) %>% ungroup() %>%
 filter(success ==1)   )  %>% select(-success , - n)  %>% left_join(play_by_play_train    %>%   filter(down >0 & goal_to_go == 1) %>% group_by(down) %>% summarise(`Mean Expected Points Added` =  round(mean(epa) , digits = 2)   , `Median Expected Points Added` = round(median(epa) , digits = 2)) ) 
```




```{r  , td_rate_over_time}
library(plotly)

p <- play_by_play_train %>%  bind_rows(play_by_play_test) %>%  filter(goal_to_go ==1) %>% filter(play_type != "no_play") %>%  count(season, down, touchdown  ) %>%  group_by(season, down) %>% mutate(`Touchdown Rate` =    round(  n / sum(n)  , digits = 3) )%>% ungroup() %>% 
  filter(touchdown ==1) %>% select(season , down , `Touchdown Rate`) %>%  ggplot(aes(x = season , y = `Touchdown Rate` ,  color = as.factor(down ))) + geom_point() +
  geom_line() + theme_tq() + scale_color_tq() + labs(title = "Goal to go TD Rate by down & year" ,  color = "Down") + scale_y_continuous(labels = scales::percent) 

p

```


```{r  , succes_rate_over_time}


p <- play_by_play_train %>%  bind_rows(play_by_play_test) %>%  filter(goal_to_go ==1) %>% filter(play_type != "no_play") %>%
  count(season, down, success  ) %>%  group_by(season, down) %>% mutate(`Success Rate` =    round(  n / sum(n)  , digits = 3) )%>% ungroup() %>% 
  filter(success ==1) %>% select(season , down , `Success Rate`) %>%  ggplot(aes(x = season , y = `Success Rate` ,  color = as.factor(down ))) + geom_point() +
  geom_line() + theme_tq() + scale_color_tq() + labs(title = "Goal to go `Success Rate` by down & year" ,  color = "Down") + scale_y_continuous(labels = scales::percent) 

p




#  no play, success vs not  this is solid


p <- play_by_play_train %>%  bind_rows(play_by_play_test) %>%  filter(goal_to_go ==1) %>% filter(play_type != "no_play") %>%
  count(season, down, success  ) %>%  group_by(season, down) %>% mutate(`Success Rate` =    round(  n / sum(n)  , digits = 3) )%>% ungroup() %>% 
  filter(success ==1) %>% select(season , down , `Success Rate`) %>%  ggplot(aes(x = season , y = `Success Rate` ,  color = as.factor(down ))) + geom_point() +
  geom_line() + theme_tq() + scale_color_tq() + labs(title = "Goal to go `Success Rate` by down & year" ,  color = "Down") + scale_y_continuous(labels = scales::percent) 

p


```






# a no play can be successful due to a penalty, but can success be predicted on no_plays?




```{r}


p <- play_by_play_train %>%  bind_rows(play_by_play_test) %>%  filter(goal_to_go ==1) %>% filter(play_type == "no_play") %>%
  count(season, down, success  ) %>%  group_by(season, down) %>% mutate(`Success Rate` =    round(  n / sum(n)  , digits = 3) )%>% ungroup() %>% 
  filter(success ==1) %>% select(season , down , `Success Rate`) %>%  ggplot(aes(x = season , y = `Success Rate` ,  color = as.factor(down ))) + geom_point() +
  geom_line() + theme_tq() + scale_color_tq() + labs(title = "Goal to go `Success Rate` by down & year" ,  color = "Down") + scale_y_continuous(labels = scales::percent) 

p

# third down benefits from penalty!

dog <-  play_by_play_train %>%  bind_rows(play_by_play_test) %>% filter(play_type == "no_play")
 
 
# %>%  #filter(goal_to_go ==1)  %>% 
 # count(down, success , play  ) %>%  group_by( down , play) %>% mutate(`Success Rate` =    round(  n / sum(n)  , digits = 3) )%>% ungroup() %>% 
  #filter(success ==1) %>% select( down , play , `Success Rate`)  %>%  gt()


 #look at the play 0 1 field
 
```




```{r}
cow_plot_epa_wpa(dog)




dog  %>%  ggplot(aes(x=wpa)) + geom_histogram() +facet_grid(~down) 
```

