---
title: 4th and Goal
author: Jake Lowell
date: '2020-04-13'
slug: 4th-and-goal
categories:
  - Football
tags:
  - NFL
description: 'In this post I explore goal to go data, and build a model to predict success on 4th and goal'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE , message = FALSE  , warning = FALSE )

#, fig.width = 8 , fig.height = 5
```



```{r  message  = FALSE}

library(gt)
library(tidyverse)
library(tidyquant)
library(glue)
library(tidymodels)
library(plotly)

# 1. Load Play by play data -----



play_by_play_2009_2012 <- readRDS( "~/Desktop/Blog/static/data/NFL_Play_by_Play_2009_2012.rds")

play_by_play_2013_2015 <- readRDS( "~/Desktop/Blog/static/data/NFL_Play_by_Play_2013_2015.rds")

play_by_play_2016_2019 <- readRDS( "~/Desktop/Blog/static/data/NFL_Play_by_Play_2016_2019.rds")


play_by_play <- play_by_play_2009_2012 %>%  bind_rows(play_by_play_2013_2015 , play_by_play_2016_2019 )

play_by_play_train <- play_by_play %>%  filter(game_date  < "2019-05-01")

play_by_play_test <- play_by_play %>%  filter(game_date  > "2019-05-01")


```








```{r    message  = FALSE}



label_wrap <- label_wrap_gen(width = 60)
#source("scripts/Load data - pass and run plays.R")

# 5  filter so that we have only run and pass data ----

pbp_all_rp <- play_by_play_train    %>%  select(  touchdown ,  td_prob , play_type , down, goal_to_go, desc , yardline_100 , yards_gained , td_team, epa, play_id, game_id, home_team, away_team, posteam, posteam_type, game_date, half_seconds_remaining , game_seconds_remaining, game_half , drive, sp, qtr, ydstogo , shotgun, no_huddle, qb_dropback, qb_scramble, pass_location, air_yards, run_location, run_gap ,posteam_timeouts_remaining , defteam_timeouts_remaining ,  ep , total_home_epa, total_away_epa, total_home_rush_epa , total_home_pass_epa , total_away_pass_epa , total_away_rush_epa ,  wp , def_wp , score_differential , posteam_score , defteam_score  , week , season ,   first_down_pass , first_down_rush , third_down_converted , fourth_down_converted  , passer_player_name , receiver_player_name  , rusher_player_name) %>% 
  filter(td_prob > 0 )%>% 
  filter(!is.na(epa), !is.na(posteam), play_type=="no_play" | play_type=="pass" | play_type=="run") %>%
  mutate(
    pass = if_else(str_detect(desc, "( pass)|(sacked)|(scramble)"), 1, 0),
    rush = if_else(str_detect(desc, "(left end)|(left tackle)|(left guard)|(up the middle)|(right guard)|(right tackle)|(right end)") & pass == 0, 1, 0),
    success = ifelse(epa>0, 1 , 0),
    passer_player_name = ifelse(play_type == "no_play" & pass == 1, 
                                str_extract(desc, "(?<=\\s)[A-Z][a-z]*\\.\\s?[A-Z][A-z]+(\\s(I{2,3})|(IV))?(?=\\s((pass)|(sack)|(scramble)))"),
                                passer_player_name),
    receiver_player_name = ifelse(play_type == "no_play" & str_detect(desc, "pass"), 
                                  str_extract(desc, "(?<=to\\s)[A-Z][a-z]*\\.\\s?[A-Z][A-z]+(\\s(I{2,3})|(IV))?"),
                                  receiver_player_name),
    rusher_player_name = ifelse(play_type == "no_play" & rush == 1, 
                                str_extract(desc, "(?<=\\s)[A-Z][a-z]*\\.\\s?[A-Z][A-z]+(\\s(I{2,3})|(IV))?(?=\\s((left end)|(left tackle)|(left guard)|	 (up the middle)|(right guard)|(right tackle)|(right end)))"),
                                rusher_player_name),
    name = ifelse(!is.na(passer_player_name), passer_player_name, rusher_player_name),
    yards_gained=ifelse(play_type=="no_play",NA,yards_gained),
    play=1  , TD_Scored = ifelse(!is.na(td_team) , "Touchdown" , "No Touchdown")  ,
    Scored_TD = case_when(td_team == posteam ~  "Touchdown" ,
                          is.na(td_team)  ~ "Not Touchdown"  , TRUE ~ "Touchdown Other Team")
  )  %>% 
  filter(play_type %in% c("run" , "pass")  | pass==1 | rush==1)  %>% 
  select(pass , td_prob  ,rush, play_type , down , goal_to_go , desc , yardline_100   , yards_gained , td_team , td_prob , epa   ,everything()) %>% 
  filter( season != 2019) %>%  mutate_if(is.character , factor ) %>% select(-game_date) %>% 
   mutate(play_desc = str_glue("Play: {label_wrap(desc)}"))
                                     

# Expected Points Added: {round(epa, digits = 2)}
#                                      Touchdown Probability: {round(td_prob, digits = 2)}
#                                      Touchdown: {Scored_TD}"))
  #mutate(play_desc =  str_glue(" Play: {label_wrap(desc)}"))
 #added play_type %in% c("run" , "pass") because this incluesd real plays where run/pass criteria is not satified in description

##  breakout by each down




first_and_goal <-  pbp_all_rp %>%  filter(down == 1 & goal_to_go ==1 ) #%>% count(touchdown)

second_and_goal <-  pbp_all_rp %>%  filter(down == 2 & goal_to_go ==1 )

third_and_goal <-  pbp_all_rp %>%  filter(down == 3 & goal_to_go ==1 )

fourth_and_goal  <-   pbp_all_rp %>%  filter(down == 4 & goal_to_go ==1 )
```




I am using data from the [nflscrapR](https://arxiv.org/abs/1802.00998) package to investigate offensive success on goal to go situations on different downs.  I will perform exploratory data analysis and modeling on regular season game data from 2009-2018, and test the performance on 2019 regular season data.



# Section 1. Initial Analysis

```{r}
# 
# four_downs <-  pbp_all_rp    %>%   filter(down >0 & goal_to_go == 1) %>% ggplot(aes( x = epa ,y = td_prob,     color = TD_Scored )) + geom_point(aes(text = play_desc)) + theme_tq() + 
#   labs(    title = glue("Expected points added by Touchdown Probability"             )  ,
#        x = "Expected Points Added" , y = "Touchdown Probability" ,  color = "" ) + scale_color_tq() +
# scale_y_continuous(labels = scales::percent)  + facet_grid(~down  , scales = "free")
# 
# ggplotly(four_downs)
```






### Figure 1.

```{r}

pbp_all_rp   %>%   filter(down >0 & goal_to_go == 1) %>% ggplot(aes( x =epa ,     fill = TD_Scored )) + geom_histogram() + theme_tq() + 
  labs(title = glue("Goal to go - Distribution of expected points added for each down"             )  ,
       x = "Expected Points Added" , y = "Plays"  ,  fill = "") + scale_color_tq() +
scale_y_continuous(labels = scales::comma)  + facet_wrap(~down   , scales = "free" )


```















If we define success as a play with a positive value for excepted points added (epa), we can see a nice split in good/bad distribution for 3rd and 4th down.  There is a decent split for second down as well.












```{r}
# pbp_all_rp %>%  filter(goal_to_go ==1) %>% filter(play_type != "no_play") %>%  count(down, touchdown  ) %>%  group_by( down) %>% mutate(`Touchdown Rate` = 
#                 round(  n / sum(n)  , digits = 2) )%>% ungroup() %>% 
#   filter(touchdown ==1) %>% select(down , `Touchdown Rate`) %>% left_join( pbp_all_rp %>%  filter(goal_to_go ==1) %>%  count(down, success  ) %>%  group_by( down) %>% mutate( `Success Rate`=    round (n / sum(n) , digits = 2)) %>% ungroup() %>%
#  filter(success ==1)   )  %>% select(-success , - n)  %>% left_join(pbp_all_rp   %>%   filter(down >0 & goal_to_go == 1) %>% group_by(down) %>% summarise(`Mean Expected Points Added` =  round(mean(epa) , digits = 2)   , `Median Expected Points Added` = round(median(epa) , digits = 2)) )   %>% DT::datatable() %>% DT::formatPercentage(c("Touchdown Rate" , "Success Rate"))
```






```{r}
pbp_all_rp %>%  filter(goal_to_go ==1) %>% filter(play_type != "no_play") %>%  count(down, touchdown  ) %>%  group_by( down) %>% mutate(`Touchdown Rate` = 
                round(  n / sum(n)  , digits = 3) )%>% ungroup() %>% 
  filter(touchdown ==1) %>% select(down , `Touchdown Rate`) %>% left_join( pbp_all_rp %>%  filter(goal_to_go ==1) %>%  count(down, success  ) %>%  group_by( down) %>% mutate( `Success Rate`=    round (n / sum(n) , digits = 3)) %>% ungroup() %>%
 filter(success ==1)   )  %>% select(-success , - n)  %>% left_join(pbp_all_rp   %>%   filter(down >0 & goal_to_go == 1) %>% group_by(down) %>% summarise(`Mean Expected Points Added` =  round(mean(epa) , digits = 2)   , `Median Expected Points Added` = round(median(epa) , digits = 2)) )   %>% gt() %>%
  fmt_percent(columns = vars(`Touchdown Rate` , `Success Rate`) , decimals = 1)  %>% tab_header(
    title = md("Goal to goal Results"),
    subtitle = ""
  ) %>%
  tab_source_note(md("Source: nflscrapR"))
```








We can hover over plays in  **Figure 2.** to see from the description of plays with penalties that penalties contribute to expected points added.  This means that a "no_play" with a penalty can be a positive or negative outcome, so I will include these plays in my "Success" models.  Since none of these plays result in a touchdown for the offence, I will exclude them from my touchdown and yardage modeling. 









### Figure 2.


```{r}

four_downs <-  pbp_all_rp  %>% filter(play_type =="no_play")  %>%   filter(down >0 & goal_to_go == 1) %>% ggplot(aes( x = epa ,y = td_prob )) + geom_point(aes(text = play_desc)) + theme_tq() + 
  labs(    title = glue("No Play -- Expected points added by Touchdown Probability"             )  ,
       x = "Expected Points Added" , y = "Touchdown Probability" , caption = "Source: nflscrapR" ) + scale_color_tq() +
scale_y_continuous(labels = scales::percent)  + facet_grid(~down  , scales = "free")

ggplotly(four_downs)
```









### Figure 3.


```{r}

four_downs <-  pbp_all_rp  %>% filter(play_type =="no_play")  %>%   filter(down >0  & goal_to_go == 1 ) %>% 
 ggplot(aes( x = epa )) + geom_histogram() + theme_tq() + 
  labs(    title = "No Play -- Distribution of Expected Points Added"            ,
       x = "Expected Points Added" , y = "Touchdown Probability" ,  color = ""    , caption = "Source: nflscrapR" ) + scale_color_tq() +
scale_y_continuous(labels = scales::comma)  + facet_wrap(~down  , scales = "free")

#,    color = TD_Scored 

four_downs

#ggplotly(four_downs)
```







# Section 2. Predict Second and goal sucess rate


Playcalling is important on second and goal.  Can we use data to help us predict offensive effectiveness on second and goal?


# Figure 4.

```{r}

second_and_goal  %>%   filter(down >0 & goal_to_go == 1) %>% ggplot(aes( x =epa ,     fill = TD_Scored )) + geom_histogram() + theme_tq() + 
  labs(title = glue("Goal to go - Distribution of expected points added for each down"             )  ,
       x = "Expected Points Added" , y = "Plays"  ,  fill = "") + scale_color_tq() +
scale_y_continuous(labels = scales::comma)  + facet_wrap(~down   , scales = "free" )

```




# Figure 5.

```{r}

second <- second_and_goal %>% filter(epa >0  & epa < 0.5) %>%   ggplot(aes( x = epa ,y = yards_gained  , color = TD_Scored)) + geom_point(aes(text = play_desc)) + theme_tq() + 
  labs(    title = glue("Second & Goal -- Expected points added by Touchdown Probability"             )  ,
       x = "Expected Points Added" , y = "Yards Gained" , caption = "Source: nflscrapR"   , color = "") + scale_color_tq() +
scale_y_continuous(labels = scales::comma)  + facet_grid(~down  , scales = "free")

ggplotly(second)
```





#  yards

If we look at plays with expected points added with a positive value of less than 0.5, I am satisfied that these are positive results.  See Figure 4.


# Figure 6.

```{r}
second_and_goal  %>%   filter(down >0 & goal_to_go == 1)  %>% ggplot(aes( x =yards_gained ,     fill = as.factor(success ))) + geom_histogram() + theme_tq() + 
  labs(title = glue("Second and Goal - Distribution of yards gained \n Colored by Success"             )  ,
       x = "Yards Gained" , y = "Plays"  ,  fill = "") + scale_color_tq() +
scale_y_continuous(labels = scales::comma)  + facet_wrap(~TD_Scored   , scales = "free" )

```



Extract insight

```{r}
library(Information)
options(scipen=999)

IV <- create_infotables(data = second_and_goal , y = "success")

knitr::kable(IV$Summary)



```

keep

first_down_pass
yardline_100
total_home_epa
total_away_epa
	play_type
		air_yards
		qb_dropback
		half_seconds_remaining
			drive
			pass_location
			run_location
			no_huddle


```{r}


#  maybe this was making the test set?    yeah, this makes 2019 only test sets
#  i don't think this was necessary.   actually maybe it was! i need to wrap up and document work in the future, moving forward.!!!!!!!!

pbp_all_test <- play_by_play_test   %>%  select(  touchdown ,  td_prob , play_type , down, goal_to_go, desc , yardline_100 , yards_gained , td_team, epa, play_id, game_id, home_team, away_team, posteam, posteam_type, game_date, half_seconds_remaining , game_seconds_remaining, game_half , drive, sp, qtr, ydstogo , shotgun, no_huddle, qb_dropback, qb_scramble, pass_location, air_yards, run_location, run_gap ,posteam_timeouts_remaining , defteam_timeouts_remaining ,  ep , total_home_epa, total_away_epa, total_home_rush_epa , total_home_pass_epa , total_away_pass_epa , total_away_rush_epa ,  wp , def_wp , score_differential , posteam_score , defteam_score  , week , season ,   first_down_pass , first_down_rush , third_down_converted , fourth_down_converted  , passer_player_name , receiver_player_name  , rusher_player_name) %>% 
  filter(td_prob > 0 )%>% 
  filter(!is.na(epa), !is.na(posteam), play_type=="no_play" | play_type=="pass" | play_type=="run") %>%
  mutate(
    pass = if_else(str_detect(desc, "( pass)|(sacked)|(scramble)"), 1, 0),
    rush = if_else(str_detect(desc, "(left end)|(left tackle)|(left guard)|(up the middle)|(right guard)|(right tackle)|(right end)") & pass == 0, 1, 0),
    success = ifelse(epa>0, 1 , 0),
    passer_player_name = ifelse(play_type == "no_play" & pass == 1, 
                                str_extract(desc, "(?<=\\s)[A-Z][a-z]*\\.\\s?[A-Z][A-z]+(\\s(I{2,3})|(IV))?(?=\\s((pass)|(sack)|(scramble)))"),
                                passer_player_name),
    receiver_player_name = ifelse(play_type == "no_play" & str_detect(desc, "pass"), 
                                  str_extract(desc, "(?<=to\\s)[A-Z][a-z]*\\.\\s?[A-Z][A-z]+(\\s(I{2,3})|(IV))?"),
                                  receiver_player_name),
    rusher_player_name = ifelse(play_type == "no_play" & rush == 1, 
                                str_extract(desc, "(?<=\\s)[A-Z][a-z]*\\.\\s?[A-Z][A-z]+(\\s(I{2,3})|(IV))?(?=\\s((left end)|(left tackle)|(left guard)|	 (up the middle)|(right guard)|(right tackle)|(right end)))"),
                                rusher_player_name),
    name = ifelse(!is.na(passer_player_name), passer_player_name, rusher_player_name),
    yards_gained=ifelse(play_type=="no_play",NA,yards_gained),
    play=1  , TD_Scored = ifelse(!is.na(td_team) , "Touchdown" , "No Touchdown")  ,
    Scored_TD = case_when(td_team == posteam ~  "Touchdown" ,
                          is.na(td_team)  ~ "Not Touchdown"  , TRUE ~ "Touchdown Other Team")
  )  %>% 
  filter(play_type %in% c("run" , "pass")  | pass==1 | rush==1)  %>% 
  select(pass , td_prob  ,rush, play_type , down , goal_to_go , desc , yardline_100   , yards_gained , td_team , td_prob , epa   ,everything())  %>%  mutate_if(is.character , factor ) %>% select(-game_date) %>% 
   mutate(play_desc = str_glue("Play: {label_wrap(desc)}"))
                                     

# Expected Points Added: {round(epa, digits = 2)}
#                                      Touchdown Probability: {round(td_prob, digits = 2)}
#                                      Touchdown: {Scored_TD}"))
  #mutate(play_desc =  str_glue(" Play: {label_wrap(desc)}"))
 #added play_type %in% c("run" , "pass") because this incluesd real plays where run/pass criteria is not satified in description

##  breakout by each down




first_and_goal_test <-  pbp_all_test %>%  filter(down == 1 & goal_to_go ==1 ) #%>% count(touchdown)

second_and_goal_test <-  pbp_all_test %>%  filter(down == 2 & goal_to_go ==1 )

third_and_goal_test <- pbp_all_test %>%  filter(down == 3 & goal_to_go ==1 )

fourth_and_goal_test  <-   pbp_all_test %>%  filter(down == 4 & goal_to_go ==1 )
```


```{r}

# second pass

library(scorecard)

train  <-  second_and_goal %>% select(
  success ,
  first_down_pass,
  yardline_100,
  total_home_epa,
  #total_away_epa,
  play_type,
  air_yards,
  qb_dropback,
  half_seconds_remaining,
  drive,
  #pass_location,
  #run_location,
  no_huddle ,first_down_pass ,
  first_down_rush ,
  half_seconds_remaining ,
  shotgun ,
  drive ,
  score_differential 
)


test  <-  second_and_goal_test %>% select(
  success ,
  first_down_pass,
  yardline_100,
  total_home_epa,
 # total_away_epa,
  play_type,
  air_yards,
  qb_dropback,
  half_seconds_remaining,
  drive,
  #pass_location,
  #run_location,
  no_huddle ,first_down_pass ,
  first_down_rush ,
  half_seconds_remaining ,
  shotgun ,
  drive ,
  score_differential 
)





filt <-  var_filter( train  , y = "success" )

bins <- woebin(filt,  y = "success" )



iv = iv(train, y = "success") %>%
  as_tibble() %>%
  mutate( info_value = round(info_value, 3) ) %>%
  arrange( desc(info_value) )

iv


bins


train_woe  <- woebin_ply(train , bins)



test_woe  <- woebin_ply(test , bins)
```




<!-- ```{r warning= FALSE , message = FALSE} -->
<!-- GGally::ggpairs( train ,aes(color =  as_factor(success))) -->
<!-- ``` -->



# correlationfunnel

```{r}
library(correlationfunnel)

jj <- cor(train_woe)

corr_table <- train_woe %>% correlate(success)  %>%  filter(feature != "success")

corr_table %>%  plot_correlation_funnel(limits = c(-0.5 , 0.5))

corr_table

```







# recipes

```{r}

library(tidymodels)

set.seed(1234)

train_woe<- train_woe %>%  mutate(Success = as_factor(success))

train_woe%>% count(Success)


test_woe <- test_woe %>%  mutate(Success = as_factor(success))

test_woe %>% count(Success)



second_rec <- recipe(Success ~ .  , data = train_woe) %>%
  #step_downsample(Success) %>%  # class imbalence 
  step_zv(all_numeric()) %>%
  prep()

second_rec 

```



```{r}
train_tbl <-  bake(second_rec , new_data = train_woe)
  
test_tbl <- bake(second_rec , new_data = test_woe)
```


```{r}
log_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

log_fit <- log_spec %>% fit(Success ~ . - success , data = train_tbl)

log_fit %>%  tidy()





tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

tree_fit <- tree_spec %>%
 fit(Success ~ . - success , data = train_tbl)

tree_fit 



```



############





```{r}

set.seed(123)
second_cv_folds <- train_tbl %>% 
    bake(second_rec, new_data = .) %>%
    vfold_cv(v = 5)


predictions_glm <- log_fit  %>%
    predict(new_data = test_tbl) %>%
    bind_cols(test_tbl %>% select(Success , success))

predictions_glm %>% head() %>% knitr::kable()


```



```{r}
predictions_glm %>%
    conf_mat(Success, .pred_class) %>%
    pluck(1) %>%
    as_tibble() %>%
    
    # Visualize with ggplot
    ggplot(aes(Prediction, Truth, alpha = n)) +
    geom_tile(show.legend = FALSE) +
    geom_text(aes(label = n), colour = "white", alpha = 1, size = 8) + labs(title= "confustion matrix")
```


```{r}
library(knitr)
predictions_glm %>%
    metrics(Success, .pred_class) %>%
    select(-.estimator) %>%
    filter(.metric == "accuracy") %>%
    kable()

test_tbl %>% count(Success)

```


<!-- ```{r} -->
<!-- log_res <- fit_resamples( -->
<!--  Success ~ ., -->
<!--   log_fit, -->
<!--   validation_splits, -->
<!--   control = control_resamples(save_pred = TRUE)) -->


<!-- knn_res %>% -->
<!--   collect_metrics() -->
<!-- ``` -->

 
 
 
 
 
 
 
# deal with later 
 

<!-- ```{r} -->
<!-- tibble( -->
<!--     "precision" =  -->
<!--         precision(predictions_glm, Success, .pred_class) %>% -->
<!--         select(.estimate), -->
<!--     "recall" =  -->
<!--         recall(predictions_glm, Success, .pred_class) %>% -->
<!--         select(.estimate) -->
<!-- ) %>% -->
<!--     unnest(cols = c(precision, recall)) %>% -->
<!--     kable() -->
<!-- ``` -->



<!-- ```{r} -->
<!-- roc_curve(predictions_glm , success , as.nu.pred_class) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- predictions_glm %>% -->
<!--     f_meas(Success, .pred_class) %>% -->
<!--     select(-.estimator) %>% -->
<!--     kable() -->
<!-- ``` -->


<!-- #  roc curve -->
<!-- ```{r} -->


<!-- log_res %>% -->
<!--   unnest(.predictions) %>% -->
<!--   mutate(model = "glm") %>% -->
<!--   bind_rows(tree_res %>% -->
<!--     unnest(.predictions) %>% -->
<!--     mutate(model = "rpart")) %>% -->
<!--   group_by(model) %>% -->
<!--   roc_curve(Success, .pred_Success) %>% -->
<!--   ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) + -->
<!--   geom_line(size = 1.5) + -->
<!--   geom_abline( -->
<!--     lty = 2, alpha = 0.5, -->
<!--     color = "gray50", -->
<!--     size = 1.2 -->
<!--   ) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- knn_conf <- knn_res %>% -->
<!--   unnest(.predictions) %>% -->
<!--   conf_mat(children, .pred_class) -->

<!-- knn_conf -->
<!-- ``` -->



<!-- ```{r} -->
<!-- knn_conf %>% -->
<!--   autoplot(type = "heatmap") + labs(title = "This is a nice little plot") -->

<!-- ``` -->

